
```{r}
save_pheatmap_png <- function(x, filename, width=1200, height=1000, res = 150) {
  png(filename, width = width, height = height, res = res)
  grid::grid.newpage()
  grid::grid.draw(x$gtable)
  dev.off()
}
```
Load necessary libraries:
```{r}
library(pheatmap)
library(dplyr)
library(xtable)
library(readr)
library(ggpubr)
```

```{r}
pubFontSize = 18
pValSize = 8
pubWidth = 8
pubHeight = 6
pubColorScheme = "Set1"
```


# Read in the prediction results and rename columns to just be model names:
```{r}
prediction_results <- read_csv("../Prediction/final_results_combined.csv")

prediction_results <- prediction_results %>%
  rename(DeepHit = DeepHit_cIndex,
         DeepSurv = DeepSurv_cIndex)
```
For readability down the line, combine the index and outcome events into one column called "Dataset" and short "account liquidated" to just "liquidated":
```{r}
prediction_results <- prediction_results %>%
  mutate(outcomeEvent = case_when(outcomeEvent == "account liquidated" ~ "liquidated",
                                  TRUE ~ outcomeEvent)) %>%
  mutate(Dataset = paste0(indexEvent, "-", outcomeEvent)) %>%
  select(-indexEvent, -outcomeEvent)
```

Rearrange the rows and columns each according to their mean Borda rank:
```{r}
# Use the Dataset column as rownames:
prediction_results <- prediction_results %>%
  column_to_rownames(var = "Dataset")

# Calculate the means of the columns (mean concordance index for each model):
col_means <- colMeans(prediction_results)
prediction_results <- bind_rows(prediction_results, col_means) 
rownames(prediction_results)[rownames(prediction_results) == "...17"] <- "Mean C-Index"

# Calculate the Borda ranks for the columns:
col_ranks <- data.frame(prediction_results) %>%
         rowwise() %>% 
         do(data.frame(t(rank(-unlist(.)))))
# This is a bit janky, but I know there are only 16 datasets, 
# so just keep the top 16 rows after the col_ranks have been computed:
col_ranks <- col_ranks %>%
  head(16)

mean_col_ranks <- colMeans(col_ranks)
prediction_results<- bind_rows(prediction_results, mean_col_ranks)
rownames(prediction_results)[rownames(prediction_results) == "...18"] <- "Mean Borda Rank"

# Order column names based on their means (ascending)
ordered_cols <- prediction_results[, order(as.numeric(prediction_results["Mean Borda Rank", ]))]

# Set up the dataframe for annotating heatmap columns based on Borda Rank
annot_cols <- as.numeric(ordered_cols["Mean Borda Rank", ])
names(annot_cols) <- colnames(ordered_cols)
annot_cols <- data.frame(`Mean Borda Rank` = annot_cols)

prediction_results_for_heatmap <- ordered_cols[1:16, ]

# Next, let's order the columns by Borda rank by transposing the matrix and doing the same thing as above:
prediction_results_transposed <- data.frame(t(prediction_results_for_heatmap)) 
row_means <- colMeans(prediction_results_transposed)
prediction_results_transposed <- bind_rows(prediction_results_transposed, row_means)
rownames(prediction_results_transposed)[rownames(prediction_results_transposed) == "...7"] <- "Mean C-Index"

row_ranks <- data.frame(prediction_results_transposed) %>%
  rowwise() %>%
  do(data.frame(t(rank(-unlist(.)))))

row_ranks <- row_ranks %>%
  head(6)

mean_row_ranks <- colMeans(row_ranks)
prediction_results_transposed <- bind_rows(prediction_results_transposed, mean_row_ranks)
rownames(prediction_results_transposed)[rownames(prediction_results_transposed) == "...8"] <- "Mean Borda Rank"

annot_rows <- as.numeric(prediction_results_transposed["Mean Borda Rank", ])
names(annot_rows) <- colnames(prediction_results_transposed)
annot_rows <- data.frame(`Mean Borda Rank` = annot_rows)

rownames(annot_rows) <- rownames(prediction_results_for_heatmap)

final_row_order <- annot_rows %>%
  arrange(Mean.Borda.Rank) %>%
  rownames()

prediction_results_for_heatmap <- prediction_results_for_heatmap[final_row_order,]


prediction_results_heatmap <- pheatmap(as.matrix(prediction_results_for_heatmap), scale='none',
         angle_col = 45,
         main = "Survival Prediction Concordance Index Values", 
         cluster_cols = FALSE, cluster_rows = FALSE,
         display_numbers = TRUE,
         annotation_col = annot_cols,
         annotation_row = annot_rows,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         width = 1280,
         height = 720
         )
```
With the heatmap created and looking good, let's save it to a PDF for publication-readiness:
```{r}
image=ggpubr::ggpar(prediction_results_heatmap,
              font.legend = list(size = pubFontSize),
              font.xtickslab = list(size = pubFontSize),
              font.ytickslab = list(size = pubFontSize),
              font.x = list(size = pubFontSize),
              font.y = list(size = pubFontSize),
              font.subtitle = list(size = pubFontSize),
              palette = pubColorScheme)
ggsave(file="predictionResults.pdf", plot=image, width=pubWidth, height=pubHeight)
```
Output the table for LaTeX in the Appendix:
```{r}
xtable(prediction_results, digits=c(0, 3, 3, 3, 3, 3, 3))
```
Now do all the same things for the classification results:
```{r}
classification_results <- read_csv("../Classification/Results/auc_score.csv")
```
Rename the auc_score column as Dataset, convert the '+' to '-', and shorten "account liquidated" to "liquidated"
```{r}
classification_results <- classification_results %>%
  rename(Dataset = auc_score) 

# Step 1: Replace " + " with "-"
classification_results$Dataset <- gsub(" \\+ ", "-", classification_results$Dataset)


# Step 2: Replace "account liquidated" with "liquidated"
classification_results$Dataset <- gsub("account liquidated", "liquidated", classification_results$Dataset, fixed = TRUE)
```

Rearrange the rows and columns according to their Mean Borda Rank:
```{r}
# Use the Dataset column as rownames:
classification_results <- classification_results %>%
  column_to_rownames(var = "Dataset")

# Also, rename "DeepLearningClassifier" as Neural Net:
classification_results <- classification_results %>%
  rename(NeuralNet = DeepLearningClassifier)

# Calculate the means of the columns (mean concordance index for each model):
col_means <- colMeans(classification_results)
classification_results <- bind_rows(classification_results, col_means) 
rownames(classification_results)[rownames(classification_results) == "...17"] <- "Mean AUC"

# Calculate the Borda ranks for the columns:
col_ranks <- data.frame(classification_results) %>%
         rowwise() %>% 
         do(data.frame(t(rank(-unlist(.)))))
# This is a bit janky, but I know there are only 16 datasets, 
# so just keep the top 16 rows after the col_ranks have been computed:
col_ranks <- col_ranks %>%
  head(16)

mean_col_ranks <- colMeans(col_ranks)
classification_results<- bind_rows(classification_results, mean_col_ranks)
rownames(classification_results)[rownames(classification_results) == "...18"] <- "Mean Borda Rank"

# Order column names based on their means (ascending)
ordered_cols <- classification_results[, order(as.numeric(classification_results["Mean Borda Rank", ]))]

# Set up the dataframe for annotating heatmap columns based on Borda Rank
annot_cols <- as.numeric(ordered_cols["Mean Borda Rank", ])
names(annot_cols) <- colnames(ordered_cols)
annot_cols <- data.frame(`Mean Borda Rank` = annot_cols)

classification_results_for_heatmap <- ordered_cols[1:16, ]

# Next, let's order the columns by Borda rank by transposing the matrix and doing the same thing as above:
classification_results_transposed <- data.frame(t(classification_results_for_heatmap)) 
row_means <- colMeans(classification_results_transposed)
classification_results_transposed <- bind_rows(classification_results_transposed, row_means)
rownames(classification_results_transposed)[rownames(classification_results_transposed) == "...7"] <- "Mean AUC"

row_ranks <- data.frame(classification_results_transposed) %>%
  rowwise() %>%
  do(data.frame(t(rank(-unlist(.)))))

row_ranks <- row_ranks %>%
  head(6)

mean_row_ranks <- colMeans(row_ranks)
classification_results_transposed <- bind_rows(classification_results_transposed, mean_row_ranks)
rownames(classification_results_transposed)[rownames(classification_results_transposed) == "...8"] <- "Mean Borda Rank"

annot_rows <- as.numeric(classification_results_transposed["Mean Borda Rank", ])
names(annot_rows) <- colnames(classification_results_transposed)
annot_rows <- data.frame(`Mean Borda Rank` = annot_rows)

rownames(annot_rows) <- rownames(classification_results_for_heatmap)

final_row_order <- annot_rows %>%
  arrange(Mean.Borda.Rank) %>%
  rownames()

classification_results_for_heatmap <- classification_results_for_heatmap[final_row_order,]


classification_results_heatmap <- pheatmap(as.matrix(classification_results_for_heatmap), scale='none',
         angle_col = 45,
         main = "Classification AUC Values", 
         cluster_cols = FALSE, cluster_rows = FALSE,
         display_numbers = TRUE,
         annotation_col = annot_cols,
         annotation_row = annot_rows,
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         width = 1280,
         height = 720
         )

```
With the heatmap created and looking good, let's save it to a PDF for publication-readiness:
```{r}
image=ggpubr::ggpar(classification_results_heatmap,
              font.legend = list(size = pubFontSize),
              font.xtickslab = list(size = pubFontSize),
              font.ytickslab = list(size = pubFontSize),
              font.x = list(size = pubFontSize),
              font.y = list(size = pubFontSize),
              font.subtitle = list(size = pubFontSize),
              palette = pubColorScheme)
ggsave(file="classificationResults.pdf", plot=image, width=pubWidth, height=pubHeight)
```
```{r}
xtable(classification_results, digits=c(0, 3, 3, 3, 3, 3, 3))
```

