---
title: "featureEvaluation"
output: html_document
---
This notebook is intended to evaluate the features created in the FinSurvival benchmark paper in order to show that most or all of them have at least some predictive value for the dataset as a whole.

The general idea for this feature evaluation is as follows: for each feature/task combination, we run a univariate cox model. From this, we calculate the p-values and come up with some way to summarize the overall results. Perhaps for each feature, we take the minimum p-value across all tasks, since that would represent the best relevancy for each feature across all tasks. Hopefully each feature would be significant for at least one task.
```{r}
# Load the necessary libraries
library(dplyr)
library(readr)
library(survival)
library(stringr)
```

```{r}
event_pairs <- list(
  c("borrow", "repay"),
  c("borrow", "deposit"),
  c("borrow", "withdraw"),
  c("borrow", "account liquidated"),
  c("repay", "borrow"),
  c("repay", "deposit"),
  c("repay", "withdraw"),
  c("repay", "account liquidated"),
  c("deposit", "borrow"),
  c("deposit", "repay"),
  c("deposit", "withdraw"),
  c("deposit", "account liquidated"),
  c("withdraw", "borrow"),
  c("withdraw", "repay"),
  c("withdraw", "deposit"),
  c("withdraw", "account liquidated")
)
```

```{r}

feature_results <- data.frame()
# Loop over each pair
for (pair in event_pairs) {
  # Extract indexEvent and outcomeEvent from the pair
  indexEvent   <- pair[1]
  outcomeEvent <- pair[2]
  
  #  Source needed scripts
  source("~/DMLR_DeFi_Survival_Benchmark/dataLoader.R")
  source("~/DMLR_DeFi_Survival_Benchmark/dataPreprocessing.R")      
  
  source("~/DMLR_DeFi_Survival_Benchmark/Prediction/cox_regression.R") 
  
  # Preprocess
  proc_list <- preprocess(train, test, usePCA = FALSE)
  train_processed <- as.data.frame(proc_list[[1]])
  test_processed  <- as.data.frame(proc_list[[2]])
  
  # Identify feature columns (exclude survival outcome + obvious non-features)
  feature_cols <- setdiff(
    colnames(train_processed),
    c("timeDiff", "status")
  )
  
  
  for (feature in feature_cols) {
  
    # Subset and drop missing values
    df <- train_processed %>%
      select(timeDiff, status, all_of(feature)) %>%
      filter(complete.cases(.))
  
    # Skip features with no variation
    if (nrow(df) == 0 || length(unique(df[[feature]])) < 2) {
      next
    }
  
    # Construct univariate Cox formula
    fml <- as.formula(
      paste0("Surv(timeDiff, status) ~ ", feature)
    )
  
    # Fit Cox model safely
    fit <- tryCatch(
      coxph(fml, data = df),
      error = function(e) NULL
    )
  
    if (is.null(fit)) {
      next
    }
  
    s <- summary(fit)
  
    # Store results
    feature_results <- rbind(
      feature_results,
      data.frame(
        indexEvent   = indexEvent,
        outcomeEvent = outcomeEvent,
        feature      = feature,
        coef         = s$coef[1, "coef"],
        hr           = s$coef[1, "exp(coef)"],
        se           = s$coef[1, "se(coef)"],
        z            = s$coef[1, "z"],
        p            = s$coef[1, "Pr(>|z|)"],
        n            = fit$n,
        n_events     = fit$nevent,
        stringsAsFactors = FALSE
      )
    )
  }
}
```
```{r}
saveRDS(feature_results, "./feature_results.rds")
```

```{r}
effect_threshold <- 0.1   # ≈ HR ≤ 0.90 or ≥ 1.11
alpha <- 0.005

feature_results <- feature_results %>%
  group_by(indexEvent, outcomeEvent) %>%
  mutate(q = p.adjust(p, method = "BH")) %>%
  ungroup()

feature_results <- feature_results %>%
  mutate(
    log_hr = log(hr),
    non_negligible = abs(log_hr) >= effect_threshold,
    detectable = (q < alpha) & non_negligible
  )

feature_summary <- feature_results %>%
  group_by(feature) %>%
  summarise(
    shows_signal = any(detectable),
    n_tasks_signal = sum(detectable),
    min_q = min(q, na.rm = TRUE),
    max_abs_log_hr = max(abs(log_hr), na.rm = TRUE),
    .groups = "drop"
  )

write_csv(feature_summary, "feature_summary.csv")
```

```{r}
# Label features by category:
time_features <- c(
  "timeOfDay",
  "dayOfWeek",
  "dayOfMonth",
  "dayOfYear",
  "quarter",
  "dayOfQuarter",
  "sinTimeOfDay",
  "cosTimeOfDay",
  "sinDayOfWeek",
  "cosDayOfWeek",
  "sinDayOfMonth",
  "cosDayOfMonth",
  "sinDayOfQuarter",
  "cosDayOfQuarter",
  "sinDayOfYear",
  "cosDayOfYear",
  "sinQuarter",
  "cosQuarter",
  "isWeekend_TRUE"
)
feature_summary <- feature_summary %>%
  mutate(
    feature_category = case_when(
      grepl("^user", feature) ~ "user",
      grepl("^market", feature) ~ "market",
      feature %in% time_features ~ "time",
      TRUE ~ "raw"
    )
  )
```

```{r}

cap_value <- quantile(
  feature_summary$max_abs_log_hr[is.finite(feature_summary$max_abs_log_hr)],
  0.99,
  na.rm = TRUE
)

feature_summary <- feature_summary %>%
  mutate(
    max_abs_log_hr_plot = pmin(max_abs_log_hr, cap_value)
  )

# Let's make a density plot of features based on their max_abs_log_hr
library(ggplot2)
plot <- ggplot(
  feature_summary,
  aes(
    x = feature_category,
    y = max_abs_log_hr_plot,
    color = feature_category
  )
) +
  geom_boxplot(width = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.35, size = 0.9) +
  geom_hline(
    yintercept = 0.1,
    linetype = "dashed",
    linewidth = 0.9,
    color = "black"
  ) +
  scale_y_sqrt() +
  labs(
    x = "Feature category",
    y = "Maximum |log(Hazard Ratio)| across tasks",
    color = "Feature category"
  ) + scale_color_brewer(palette = "Set2") +
  theme_minimal() +
  theme(
    legend.position = "none"
  )


plot

```

```{r}
ggsave("featureEvalPlot.pdf", plot)
```